---
title: "Animated plots of spectral data"
subtitle: "'ggspectra' `r packageVersion('ggspectra')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Animated plots of spectral data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

_This article is only available as part of [the on-line documentation of R package 'ggspectra'](https://docs.r4photobiology.info/ggspectra/). It is not installed in the local computer as part of the package._

Package `ggspectra` extends `ggplot2` with stats, geoms, scales and annotations suitable for light- or radiation-related spectra. It also defines `ggplot()` and `autoplot()` methods specialized for the classes defined in package `photobiology` for storing different types of spectral data. The `autoplot()` methods are described separately in vignette _User Guide: 2 Autoplot Methods_ and the `ggplot()` methods, statistics, and scales in _User Guide: 1 Grammar of Graphics_.

The new elements can be freely combined with methods and functions defined in packages 'ggplot2', `scales`, `ggrepel`, `gganimate` and other extensions to 'ggplot2'.

This article, focuses on plots of time-series of spectra created with R package
'ggspectra' and animated with package ['gganimate'](https://gganimate.com/).
Time series of spectra stored in one of the classes from R package
['photobiology'](https://docs.r4photobiology.info/photobiology/) can be acquired
or imported easily when using Ocean Optics spectrometers, with R package
['ooacquire'](https://docs.r4photobiology.info/ooacquire/). Such data can also
be obtained with radiation transfer models u other brands of spectrometers and
in many cases imported with R package
['photobiologyInOut](https://docs.r4photobiology.info/photobiologyInOut/).
Similarly, kinetics of chemical reactions involving light absorbing reactants
and/or products as well as any other temporal change in colour and or intensity
can be characterised through spectral data. In addition to time, as exemplified
below using time-series, other series of measurements in response to changes in
other continuous explanatory variables are well suited to animations.

_One limitation of the current version of 'gganimate' is that it does not
support scales names that are R _expressions_ as returned by R functions 
`parse()` and `expression()`. Functions `str2lang()` and `bquote()` have to
be used instead. These functions are used within 'ggspectra' (>= 0.3.16) 
improving compatibility._

**A difficulty remains in that `unit.out` by default is in layer functions
that of an R option. This option does not seem to be in all cases available
when the animation is assembled.**

## Set up 

```{r}
library(ggplot2)
library(photobiology)
library(photobiologySun)
library(photobiologyWavebands)
library(ggspectra)
energy_as_default()

library(lubridate)
# if suggested packages are available
gganimate_installed <- requireNamespace("gganimate", quietly = TRUE)
eval_chunks <- gganimate_installed
if (eval_chunks) {
  library(gganimate)
  message("'gganimate' version ", format(packageVersion('gganimate')), " loaded and attached.")
} else {
  message("Please, install package 'gganimate'.")
}
```

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE)
options(warnPartialMatchArgs = FALSE)
```

We change the default theme.

```{r}
theme_set(theme_bw())
```

## A short time series of spectra

We change the labels in the factor so that the time of measurement can be shown.

```{r}
my_sun.spct <- sun_evening.spct
time <- when_measured(my_sun.spct, as.df = TRUE)
my_sun.spct$spct.idx <- factor(my_sun.spct$spct.idx, 
                               levels = time[["spct.idx"]], 
                               labels = as.character(round_date(time[["when.measured"]], unit = "minute")))
my_sun.spct
```

We create a base plot to play with:

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  aes(linetype = spct.idx) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

We can animate a plot in different ways. Here, using long transitions between spectra. During the transitions interpolated data values are used to create an artificial smooth transition.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```
The default base of expression for irradiance and spectral irradiance is `"energy"`, while `"photon"` (= quantum) is also supported. Below we edited the plot code immediately above adding `unit.out = "photon"` to `ggplot()` and replacing _y_ scale `scale_y_s.e.irrad_continuous()` with `scale_y_s.q.irrad_continuous()`.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct, unit.out = "photon") +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.q.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```

Next, we add an indication of which spectrum is being displayed. Here we define the title and subtitle of the plot to be dynamic, showing information about the currently displayed frame of the animation.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```

It is possible to add some annotations. Here energy irradiances for red and far-red light.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct) +
  geom_line() +
  stat_wb_label(aes(colour = after_stat(wb.color)),
                w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                ypos.fixed = 0.082, by.group = TRUE) +
  stat_wb_e_irrad(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  ypos.fixed = 0.078) +
  stat_wb_hbar(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro"))) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  scale_colour_identity() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```
The code above edited to use photons as base of expression by changing as above, but also adjusting the position of annotations and their multiplier. While the scale and annotations are expresses in $\mu\mathrm{mol}\,\mathrm{m}^{-2}\,\mathrm{s}^{-1}\,\mathrm{nm}^{-1}$ the data are stored as $\mathrm{mol}\,\mathrm{m}^{-2}\,\mathrm{s}^{-1}\,\mathrm{nm}^{-1}$.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct, unit.out = "photon") +
  geom_line() +
  stat_wb_label(aes(colour = after_stat(wb.color)),
                w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                ypos.fixed = 0.37e-6, by.group = TRUE) +
  stat_wb_q_irrad(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  label.mult = 1e6, 
                  ypos.fixed = 0.35e-6) +
  stat_wb_hbar(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro"))) +
  scale_x_wl_continuous() +
  scale_y_s.q.irrad_continuous() +
  scale_colour_identity() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```


## A longer time series of spectra

```{r}
my_sun.spct <- sun_hourly_august.spct
```

Object `my_sun.spct` contains multiple spectra in long form. Variable `UTC` identifies which rows belong to each spectrum, based on date and time. The data cover most of the daytime of two successive days. We select data for one day.

```{r}
my_sun.spct <- subset(my_sun.spct, day(UTC) == 21)
summary(my_sun.spct)
```


```{r}
idfactor <- getIdFactor(my_sun.spct)
idfactor
class(my_sun.spct[[idfactor]])
unique(my_sun.spct[[idfactor]])
```

For building an animation, we need the indexing variable to be a `factor`.

```{r, eval=eval_chunks, warning=FALSE}
my_sun.spct[["time"]] <- my_sun.spct[[idfactor]]
my_sun.spct[[idfactor]] <- factor(my_sun.spct[[idfactor]])
length(levels(my_sun.spct[[idfactor]]))
```

We can display all 16 spectra in a single static plot.

```{r, eval=eval_chunks, warning=FALSE}
ggplot(data = my_sun.spct) +
  aes(group = get(idfactor)) +
  geom_line(alpha = 1/3, na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

The same data as above, but displayed as an animated plot.

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = my_sun.spct) +
  geom_line(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

To more easily see the differences in the shape of the spectra we can scale them to an equal total irradiance, i.e., equal area under the curves. To emphasize this, we can use `geom_spct()` instead of `geom_line()` that was used in the examples above 

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = e_irrad, target = 100),
               unit.out = "energy") +
  geom_spct(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous(name = s.e.irrad_label(scaled = TRUE)) +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = q_irrad, target = 100),
               unit.out = "photon") +
  geom_spct(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous(name = s.e.irrad_label(scaled = TRUE)) +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

## `autoplot()` partial support

It is possible to animate a ggplot created with an `autoplot()` method. Currently
`by.group = TRUE` has to be added to the call. This is because annotations
need to be added by group instead of by panel, so that they are present in
all frames.

All label definitions in 'ggspectra' (>= 0.3.16) are static or defined using
`bquote()` as required by 'gganimate'. 

```{r}
autoplot(my_sun.spct, by.group = TRUE) +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```

**Currently, `label.qty = "total"` and `label.qty = "irrad"` trigger an ERROR for photon units when attempting to animate a plot created with `autoplot()`.**

```{r, eval=FALSE}
# broken!!
photon_as_default()
autoplot(my_sun.spct, by.group = TRUE) +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```


```{r}
# works!!
photon_as_default()
autoplot(my_sun.spct, by.group = TRUE, label.qty = "sirrad", 
         unit.out = "photon") +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```
