---
title: "Animated plots of spectral data"
subtitle: "'ggspectra' `r packageVersion('ggspectra')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Animated plots of spectral data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

_This article is only available as part of [the on-line documentation of R package 'ggspectra'](https://docs.r4photobiology.info/ggspectra/). It is not installed in the local computer as part of the package._

Package `ggspectra` extends `ggplot2` with stats, geoms, scales and annotations suitable for light- or radiation-related spectra. It also defines `ggplot()` and `autoplot()` methods specialized for the classes defined in package `photobiology` for storing different types of spectral data. The `autoplot()` methods are described separately in vignette _User Guide: 2 Autoplot Methods_ and the `ggplot()` methods, statistics, and scales in _User Guide: 1 Grammar of Graphics_.

The new elements can be freely combined with methods and functions defined in packages 'ggplot2', `scales`, `ggrepel`, `gganimate` and other extensions to 'ggplot2'.

This article, focuses on plots of time-series of spectra created with R package
'ggspectra' and animated with package ['gganimate'](https://gganimate.com/).
Time series of spectra stored in one of the classes from R package
['photobiology'](https://docs.r4photobiology.info/photobiology/) can be acquired
or imported easily when using Ocean Optics spectrometers, with R package
['ooacquire'](https://docs.r4photobiology.info/ooacquire/). Such data can also
be obtained with radiation transfer models or other brands of spectrometers and
in many cases imported with R package
['photobiologyInOut](https://docs.r4photobiology.info/photobiologyInOut/).
Similarly, kinetics of chemical reactions involving light absorbing reactants
and/or products as well as any other temporal change in colour and or intensity
can be characterised through spectral data. In addition to time, as exemplified
below using time-series, other series of measurements in response to changes in
other continuous explanatory variables are well suited to animations.

_One limitation of the current version of 'gganimate' is that it does not
support scales names that are R _expressions_ as returned by R functions 
`parse()` and `expression()`. Functions `str2lang()` and `bquote()` have to
be used instead. These functions are used within 'ggspectra' (>= 0.3.16) 
improving compatibility . _

'gganimate' imposes a couple of restrictions: a) statistics must generate output for each group in groupings created by mapping aesthetics to factors, as image frames in animations behave like groups, b) expressions used as labels when including substitution of values from data must be defined using `bquote()` so that the subtitution takes place at the correct stage in the building of the animations. Deviations from these requirements made 'ggspectra' < 3.16.0 incompatible with 'gganimate'. These two requirements are now fulfilled.

**A difficulty remains in that within 'ggspectra' and 'photobiology' the default argument for `unit.out` is determined through an R option. When not set, `"energy"` is used. This option does not seem to be in all cases visible when an animation is assembled leading to inconsistencies that break the animation code when the R option is set to `"photon"` instead of `"energy"`. Currently, the solution is to pass `unit.out = "photon"` explicitly in the function calls when creating animations.**

**Compatibility with 'gganimate' is rather fragile, and may break when 'ggplot2', 'gganimate' or R itself are updated. Thus, long-term compatibility between 'ggspectra' and 'gganimate' cannot be assured.**

## Set up 

```{r}
library(ggplot2)
library(photobiology)
library(photobiologySun)
library(photobiologyWavebands)
library(ggspectra)
energy_as_default()

library(lubridate)
# if suggested packages are available
gganimate_installed <- requireNamespace("gganimate", quietly = TRUE)
eval_chunks <- gganimate_installed
if (eval_chunks) {
  library(gganimate)
  message("'gganimate' version ", format(packageVersion('gganimate')), " loaded and attached.")
} else {
  message("Please, install package 'gganimate'.")
}
```

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE)
options(warnPartialMatchArgs = FALSE)
```

We change the default theme.

```{r}
theme_set(theme_bw())
```

## Using the grammar of graphics

### A time series of five spectra

We change the labels in the factor so that the time of measurement can be shown.

```{r}
my_sun.spct <- sun_evening.spct
time <- when_measured(my_sun.spct, as.df = TRUE)
my_sun.spct$spct.idx <- factor(my_sun.spct$spct.idx, 
                               levels = time[["spct.idx"]], 
                               labels = as.character(round_date(time[["when.measured"]], unit = "minute")))
summary(my_sun.spct)
```

We create a base plot to play with:

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  aes(linetype = spct.idx) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

We can animate a plot in different ways. Here, using long transitions between spectra. During the transitions interpolated data values are used to create an artificial smooth transition.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```
The default base of expression for irradiance and spectral irradiance is `"energy"`, while `"photon"` (= quantum) is also supported. Below we edited the plot code immediately above adding `unit.out = "photon"` to `ggplot()` and replacing _y_ scale `scale_y_s.e.irrad_continuous()` with `scale_y_s.q.irrad_continuous()`.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct, unit.out = "photon") +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.q.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```

Next, we add an indication of which spectrum is being displayed. Here we define the title and subtitle of the plot to be dynamic, showing information about the currently displayed frame of the animation.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```

Some statistics have compute functions that work by ggplot panel instead of by
ggplot group to avoid redundant output in regular plots. To be used in
animations, statistics should generate output for each group, as groups are the
basis of the frames in the animation. Starting from 'ggspectra' 3.16.0
statistics that normally produce a single output per panel have a formal
parameter `by.group` that when passed `TRUE` changes their behaviour to
repeatedly produce the same output for each group. A typical example are
"annotation-like" output, like labels for wavebands with `stat_wb_label()`.
These statistics use the data mostly as reference for positioning and extent,
but not directly. For example, labels for wavebands located automatically just
above the top of the tallest spectrum in a normal figure panel and "trimmed" to
the wavelengths in data. In an animation, they must be added, instead of once
per plot to each individual image frame, that internally are ggplot groups,
rather than panels.

Thus, it is possible to add various annotations. Here energy irradiances for red and far-red light together with labels for the wavebands.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct) +
  geom_line() +
  stat_wb_label(aes(colour = after_stat(wb.color)),
                w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                ypos.fixed = 0.082, 
                by.group = TRUE) + # needed for animation
  stat_wb_e_irrad(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  ypos.fixed = 0.078) +
  stat_wb_hbar(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro"))) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  scale_colour_identity() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```
The code above edited to use photons as base of expression by changing as above, but also adjusting the position of annotations and their multiplier. While the scale and annotations are expresses in $\mu\mathrm{mol}\,\mathrm{m}^{-2}\,\mathrm{s}^{-1}\,\mathrm{nm}^{-1}$ the data are stored as $\mathrm{mol}\,\mathrm{m}^{-2}\,\mathrm{s}^{-1}\,\mathrm{nm}^{-1}$.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct, unit.out = "photon") +
  geom_line() +
  stat_wb_label(aes(colour = after_stat(wb.color)),
                w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                ypos.fixed = 0.37e-6, by.group = TRUE) +
  stat_wb_q_irrad(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  label.mult = 1e6, 
                  ypos.fixed = 0.35e-6) +
  stat_wb_hbar(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro"))) +
  scale_x_wl_continuous() +
  scale_y_s.q.irrad_continuous() +
  scale_colour_identity() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```

### A time series of 15 spectra

_With these examples `animate()` crashed R 4.5.0 with 'ggplot2' 3.5.2 and 'gganimate' 1.0.9. I do not know where the trouble originated, however setting the factor describing the times differently in the spectral data, seems to have made the examples work once again. There is still one border line example that fails._

Object `sun_hourly_august.spct` contains multiple spectra in long form. Variable `UTC` indicate the date and time for each spectrum, and factor `spct.idx` identifies the spectra using a name. The data included in package 'photobiologySun' cover most of the daytime of two successive days. We select data for one day.

```{r}
# select data for 21 August
my_sun.spct <- 
  subset(sun_hourly_august.spct, 
         day(UTC) == 21 & spct.idx != "spct.30")
# number of spectra
getMultipleWl(my_sun.spct)
# levels of the factor identifying the individual spectra
levels(my_sun.spct[[getIdFactor(my_sun.spct)]])
# summary of spectra
summary(my_sun.spct, expand = "collection")
```

We replace the indexing factor with one using local time of day for levels
instead of arbitrary names. This makes it easier to display the times in the
animations.

```{r}
# replace the values used for indexing to times of day
my_sun.spct[[getIdFactor(my_sun.spct)]] <- 
  factor(format.POSIXct(my_sun.spct[["UTC"]],
                 format = "%H:%M",
                 tz = "Europe/Helsinki"))
# rename the factor used for indexing
setIdFactor(my_sun.spct, "tod")
getIdFactor(my_sun.spct)
summary(my_sun.spct[[getIdFactor(my_sun.spct)]])
# head
my_sun.spct
# summary of spectra
summary(my_sun.spct, expand = "collection")
```

All 15 spectra in a single static plot, without using 'gganimate'. It is not easy to get a sense of how the spectrum varies through the day.

```{r}
ggplot(data = my_sun.spct) +
  aes(group = tod) +
  geom_line(alpha = 1/3, na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

The same data as above, but displayed as an animated plot. This plot communicates better the temporal dynamics of spectral irradiance.

```{r, eval=eval_chunks, warning=FALSE}
getIdFactor(my_sun.spct)
anim <- ggplot(data = my_sun.spct) +
  geom_line(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} local time (EEST)',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

To more easily see the differences in the shape of the spectra we can scale them to an equal total irradiance, $100\,\mathrm{W\,m^{-2}}$, i.e., equal area under each of the curves. To visually emphasize this, we use `geom_spct()` instead of `geom_line()` used in the examples above.

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = e_irrad, target = 100),
               unit.out = "energy") +
  geom_spct(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous(name = s.e.irrad_label(scaled = TRUE)) +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

A similar plot as above can be created expressing the spectral irradiance in photon-based units instead of in energy-based units. We scale the spectra to $1\,000\,\mathrm{\mu mol\,m^{-2}\,s{-1}}$.

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = q_irrad, target = 1e-3),
               unit.out = "photon") +
  geom_spct(na.rm = TRUE) +
  scale_x_wl_continuous() +
  scale_y_s.q.irrad_continuous(name = s.q.irrad_label(scaled = TRUE)) +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

## `autoplot()` partial support

It is possible to animate ggplots created with `autoplot()` methods. However, because in animations "plot decorations" need to be added by group instead of by panel, `by.group = TRUE` has to be added to the call, which is otherwise as for a static plot. (Internally `by.group = TRUE` is passed to the statistics from 'ggspectra' that support it, as described above.).

All label definitions created by functions and methods from 'ggspectra' (>= 0.3.16) are static or defined using `bquote()` as required by 'gganimate'.

Currently there is a further limitation: _changing the default for `unit.out` into `"photon"` using the R option results in an error unless it is also passed as argument in the call to `autoplot()`._

```{r, eval=eval_chunks}
autoplot(my_sun.spct, by.group = TRUE) +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1)
```

The conversion to photon based units works as expected when set in the call.

```{r, eval=eval_chunks}
autoplot(my_sun.spct, by.group = TRUE, label.qty = "sirrad", 
         unit.out = "photon") +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1)
```

**Currently, when `unit.out` is set through an R option but not in the call, an error is triggered when the plot is animated.**

```{r, eval=FALSE}
# broken!!
photon_as_default()
autoplot(my_sun.spct, by.group = TRUE) +
  transition_states(tod,
                    transition_length = 2,
                    state_length = 1)
```
