---
title: "Article: Animated plots of spectral data"
subtitle: "'ggspectra' `r packageVersion('ggspectra')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Article: Animated plots of spectral data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Package `ggspectra` extends `ggplot2` with stats, geoms, scales and annotations suitable for light- or radiation-related spectra. It also defines `ggplot()` and  `autoplot()` methods specialized for the classes defined in package `photobiology` for storing different types of spectral data. The `autoplot()` methods are described separately in vignette _User Guide: 2 Autoplot Methods_ and the `ggplot()` methods, statistics, and scales in _User Guide: 1 Grammar of Graphics_.

The new elements can be freely combined with methods and functions defined in packages 'ggplot2', `scales`, `ggrepel`, `gganimate` and other extensions to 'ggplot2'.

This article, focuses on the animation of plots of time-series of spectra, which can be acquired easily using Ocean Optics spectrometers toether with R package 'ooacquire'. Such data can also be obtained with radiation transfer models. Simlarly, kinetics of chemical reactions involving light absorbing reactants and/or products as well as any other temporal change in colour and or intensity can be characterised through spectral data. In addition to time, as exemplified below using time-series, other series of measurements in response to changes in other continuous explanatory variables are well suited to animations. 

Currently, plots of spectra created with method `autoplot()` cannot be animated with 'gganimate'.

## Set up 

```{r}
library(ggplot2)
library(photobiology)
library(photobiologySun)
library(photobiologyWavebands)
library(ggspectra)
library(lubridate)
# if suggested packages are available
gganimate_installed <- requireNamespace("gganimate", quietly = TRUE)
eval_chunks <- gganimate_installed
if (eval_chunks) {
  library(gganimate)
} else {
  message("Please, install package 'gganimate'.")
}
```

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE)
options(warnPartialMatchArgs = FALSE)
```

We change the default theme.

```{r}
theme_set(theme_bw())
```

## A short time series of spectra

We change the labels in the factor so that the time of measurement can be shown.

```{r}
my_sun.spct <- sun_evening.spct
time <- when_measured(my_sun.spct, as.df = TRUE)
my_sun.spct$spct.idx <- factor(my_sun.spct$spct.idx, 
                               levels = time[["spct.idx"]], 
                               labels = as.character(round_date(time[["when.measured"]], unit = "minute")))
my_sun.spct
```

We create a base plot to play with:

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  aes(linetype = spct.idx) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

We can animate a plot in different ways. Here, using long transitions between spectra. During the transitions interpolated data values are used to create an artificial smooth transition.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```

We add an indication of which spectrum is being displayed. Here we define the title and subtitle of the plot to be dynamic, showing information about the currently displayed frame of the animation.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```

It is possible to add some annotations. Here energy irradiances for red and far-red light.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct |> smooth_spct(method = "supsmu", strength = 0.5)) +
  geom_line() +
  stat_wb_e_irrad(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  ypos.fixed = 0.078) +
  stat_wb_hbar(aes(colour = after_stat(wb.color)),
                  w.band = list(Red("Sellaro"), Far_red("Sellaro")),
                  ypos.fixed = 0.082) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  scale_colour_identity() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
```


## A longer time series of spectra

```{r}
my_sun.spct <- sun_hourly_august.spct
my_sun.spct
```

Object `my_sun.spct` contains multiple spectra in long form. A variable identifies which rows belong to each spectrum, based on date and time. The data cover most of the daytime of two successive days. We select data for one day.

```{r}
my_sun.spct <- subset(my_sun.spct, day(UTC) == 21)
```


```{r}
idfactor <- getIdFactor(my_sun.spct)
idfactor
class(my_sun.spct[[idfactor]])
unique(my_sun.spct[[idfactor]])
```

For building an animation, we need the indexing variable to be a `factor`.

```{r, eval=eval_chunks, warning=FALSE}
my_sun.spct[["time"]] <- my_sun.spct[[idfactor]]
my_sun.spct[[idfactor]] <- factor(my_sun.spct[[idfactor]])
length(levels(my_sun.spct[[idfactor]]))
```

We can display all 16 spectra in a single static plot.

```{r, eval=eval_chunks, warning=FALSE}
ggplot(data = my_sun.spct) +
  aes(group = get(idfactor)) +
  geom_line(alpha = 1/3) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

The same data as above, but displayed as an animated plot.

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

To more easily see the differences in the shape of the spectra we can scale them to an equal total irradiance, i.e., equal area under the curves. To emphasize this, we can use `geom_spct()` instead of `geom_line()` that was used in the examples above 

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = e_irrad, target = 100),
               unit.out = "energy") +
  geom_spct() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous(name = s.e.irrad_label(scaled = TRUE)) +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

```{r, eval=eval_chunks, warning=FALSE}
anim <- ggplot(data = fscale(my_sun.spct, f = q_irrad, target = 100),
               unit.out = "photon") +
  geom_spct() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous(name = s.e.irrad_label(scaled = TRUE)) +
  transition_states(get(idfactor),
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state} UTC',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 20, fps = 10)
```

## `autoplot()` partial support

It is possible to animate a ggplot created with an `autoplot()` method. Currently
`by.group = TRUE` has to be added to the call. This is because annotations
need to be added by group instead of by panel, so that they are present in
each frame.

More problematic is the bug that fails to render the R expressions in the
axis labels and instead copies them as text. Some bad interaction between
'ggspectra' and 'gganimate' seems to be the cause.

```{r}
autoplot(my_sun.spct, by.group = TRUE, annotations = "", unit.out = "energy") +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```


```{r, eval=FALSE}
# broken!!
autoplot(my_sun.spct, by.group = TRUE, annotations = "", unit.out = "photon") +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```
