---
title: "Article: Animated plots of spectral data"
subtitle: "'ggspectra' `r packageVersion('ggspectra')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Article: Plotting transformed spectral data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Package `ggspectra` extends `ggplot2` with stats, geoms, scales and annotations suitable for light spectra. It also defines `ggplot()` and  `autoplot()` methods specialized for the classes defined in package `photobiology` for storing different types of spectral data. The `autoplot()` methods are described separately in vignette _User Guide: 2 Autoplot Methods_ and the `ggplot()` methods, statistics, and scales in _User Guide: 1 Grammar of Graphics_.

The new elements can be freely combined with methods and functions defined in packages 'ggplot2', `scales`, `ggrepel`, `gganimate` and other extensions to 'ggplot2'.

This article, focuses on the animation of plots of time-series of spectra, which can be obtained easily with the current version of package 'ooacquire' from Ocean Optics spectrometers. Such data can also be obtained with radiation transfer models. Kinetics of chemical reactions, or any other temporal change in colour and or intensity. In addition to time, other continuous explanatory variables are well suited to using animations. 

Currently, plots of spectra created with method `autoplot()` cannot be animated with 'gganimate'.

## Set up 

```{r}
library(ggplot2)
library(photobiology)
library(photobiologySun)
library(photobiologyWavebands)
library(ggspectra)
library(lubridate)
# if suggested packages are available
gganimate_installed <- requireNamespace("gganimate", quietly = TRUE)
eval_chunks <- gganimate_installed
if (eval_chunks) {
  library(gganimate)
} else {
  message("Please, install package 'gganimate'.")
}
```

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE)
options(warnPartialMatchArgs = FALSE)
```

We change the default theme.

```{r}
theme_set(theme_bw())
```

## Short series

We change the labels in the factor so that the time of measurement can be shown.

```{r}
my_sun.spct <- sun_evening.spct
time <- when_measured(my_sun.spct, as.df = TRUE)
my_sun.spct$spct.idx <- factor(my_sun.spct$spct.idx, 
                               levels = time[["spct.idx"]], 
                               labels = as.character(round_date(time[["when.measured"]], unit = "minute")))
my_sun.spct
```

We create a base plot to play with:

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  aes(linetype = spct.idx) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```

We can animate it in different ways.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1)
```
We need some indication of what spectrum is being displayed.

```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state}',
          subtitle = 'Frame {frame} of {nframes}')
```

It is possible to add some annotations. Here some energy irradiances.

```{r, eval=eval_chunks}
ggplot(data = sun_evening.mspct |> smooth_spct(method = "supsmu", strength = 0.5)) +
  geom_line() +
  stat_wb_e_irrad(w.band = list(PAR(), Red("Sellaro"), Far_red("Sellaro")), ypos.fixed = 0.075) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1,
                    ) +
  ggtitle('Now showing {closest_state}',
          subtitle = 'Frame {frame} of {nframes}')
```


## A longer time series

```{r}
my_sun.spct <- sun_hourly_august.spct
time <- when_measured(my_sun.spct, as.df = TRUE)
my_sun.spct$spct.idx <- factor(my_sun.spct$spct.idx, 
                               levels = time[["spct.idx"]], 
                               labels = as.character(round_date(time[["when.measured"]], unit = "hour")))
my_sun.spct
```
```{r, eval=eval_chunks}
ggplot(data = my_sun.spct) +
  aes(group = spct.idx) +
  geom_line(alpha = 1/3) +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous()
```


```{r, eval=eval_chunks}
anim <- ggplot(data = my_sun.spct) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state}',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 40, fps = 10)
```

```{r, eval=eval_chunks}
anim <- ggplot(data = fscale(my_sun.spct, target = 0.5)) +
  geom_line() +
  scale_x_wl_continuous() +
  scale_y_s.e.irrad_continuous() +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state}',
          subtitle = 'Frame {frame} of {nframes}')
animate(anim, duration = 40, fps = 10)
```
