---
title: "Plotting transformed spectral data"
subtitle: "'ggspectra' `r packageVersion('ggspectra')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Plotting transformed spectral data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

_This article is only available as part of [the on-line documentation of R package 'ggspectra'](https://docs.r4photobiology.info/ggspectra/). It is not installed in the local computer as part of the package._

Package `ggspectra` extends `ggplot2` with stats, geoms, scales and annotations suitable for light spectra. It also defines `ggplot()` and  `autoplot()` methods specialized for the classes defined in package `photobiology` for storing different types of spectral data. The `autoplot()` methods are described separately in vignette _User Guide: 2 Autoplot Methods_ and the `ggplot()` methods, statistics, and scales in _User Guide: 1 Grammar of Graphics_.

The new elements can be freely combined with methods and functions defined in packages 'ggplot2', `scales`, `ggrepel`, `gganimate`, `cowplot` and other extensions to 'ggplot2'.

This article, formerly the third part of the User Guide, describes how to combine manipulation of spectral data with plotting. This streamlined coding is made possible by an enhancement implemented in 'ggspectra' (>= 0.3.5). In addition, some of the examples make use of methods available only in 'photobiology' (>= 0.10.0).

In 'ggspectra' (>= 0.3.5) the data member of `gg` (ggplot) objects remains as an object of the classes for spectral data defined in 'photobiology' instead of being converted into a plain `data.frame`. This makes it possible data manipulations in layers to be done with methods specific to spectral data. In other words, the `data` object used as default for individual plot layers retains its original attributes, including its class. This makes it possible to use methods applicable to the original object to modify it for individual plot layers.

The examples in this vignette depend conditionally on packages 'rlang', 'magrittr', and 'gganimate'. If these packages are not available when the article is built, the code chunks that require them are not evaluated.

## Set up 

```{r}
library(ggplot2)
library(photobiology)
library(photobiologyWavebands)
library(ggspectra)
# if suggested packages are available
magrittr_installed <- requireNamespace("magrittr", quietly = TRUE)
rlang_installed <- requireNamespace("rlang", quietly = TRUE)
eval_chunks <- magrittr_installed && rlang_installed
if (eval_chunks) {
  library(magrittr)
  library(rlang)
} else {
  message("Please, install packages 'rlang' and 'magrittr'.")
}
eval_gganimate <- requireNamespace("gganimate", quietly = TRUE)
if (eval_gganimate) {
library(gganimate)
message("'gganimate' version ", 
        format(packageVersion('gganimate')), 
        " loaded and attached.")
}
```

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4,
               cache = FALSE)
options(warnPartialMatchArgs = FALSE)
```

We change the default theme.

```{r}
theme_set(theme_bw())
```

## Applying methods to spectral data in plot layers

### A single spectrum

In 'ggspectra' (< 0.3.5) we had to pass to the `data` parameter of layer functions always a data frame, or a transformation based on a method with an specialization for `data,frame`. A simple example passing spectral data to each layer function follows. Here to be able to use a method defined for `source_spct` objects like `sun.spct` we pass `sun.spct` as argument to method `smooth_spct()`.

```{r, eval=eval_chunks}
ggplot() + 
  geom_line(data = sun.spct, mapping = aes(w.length, s.e.irrad)) + 
  geom_line(data = smooth_spct(sun.spct, method = "supsmu"), 
            mapping = aes(w.length, s.e.irrad), 
            colour = "red", linewidth = 1)
```

In 'ggspectra' (< 0.3.5) we could also use R's own pipe operator to make it easier to see the intention of the code, but still had to supply `sun.spct` twice.

```{r, eval=eval_chunks}
ggplot() + 
  geom_line(data = sun.spct, mapping = aes(w.length, s.e.irrad)) + 
  geom_line(data = sun.spct |> smooth_spct(method = "supsmu"), 
            mapping = aes(w.length, s.e.irrad), 
            colour = "red", linewidth = 1)
```

In 'ggspectra' (>= 0.3.5) the class of the spectral objects stored by calls to `ggplot()` methods specific
to them is not stripped, neither are other attributes used by package
'photobiology'. Consequently, transformations in layers using default `data` can
use the specialized methods from package 'photobiology'. The next two code chunks query the class and print a summary to demonstrate this.

```{r, eval = eval_chunks}
p <- ggplot(sun.spct) + geom_line()
class(p$data)
```

```{r, eval = eval_chunks}
summary(p$data)
```

To ensure that the same data are used in both plot layers the code can be simplified using `.` to refer to the default data in the ggplot object (`p$data` in the example above). The mapping to aesthetics remains valid because `smooth_spct()` returns a new `source_spct` object with the same column names as `sun.spct`, which was passed as argument to `data`.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line() + 
  geom_line(data = . %>% smooth_spct(method = "supsmu"), 
            colour = "red", linewidth = 1)
```

Alternatively, without depending on 'magrittr' or 'rlang' we can pass an anonymous function as argument to `data` to the same effect, but possibly less clear code.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line() + 
  geom_line(data = function(x) {smooth_spct(x, method = "supsmu")}, 
            colour = "red", linewidth = 1)
```

The easiest approach to plotting photon spectral irradiance instead of spectral energy irradiance is to temporarily change the default radiation unit. An alternative approach is to replace the first two lines in the code chunk below by: `ggplot(sun.spct, unit.out = "photon") +`.

```{r, eval = eval_chunks}
photon_as_default()
ggplot(sun.spct) + 
  geom_line() + 
  geom_line(data = . %>% smooth_spct(method = "supsmu"), 
            colour = "red", linewidth = 1)
unset_radiation_unit_default()
```

Obviously, the default plot data does not need to be plotted, so this provides a roundabout way of applying methods,

```{r, eval=FALSE}
ggplot(sun.spct) + 
  geom_line(data = . %>% smooth_spct(method = "supsmu"), 
            colour = "red", linewidth = 1)
```
 
which is equivalent to doing the transformation ahead of plotting.
 
```{r, eval=eval_chunks}
sun.spct |> 
  smooth_spct(method = "supsmu") |>
  ggplot() + 
  geom_line(colour = "red", linewidth = 1)
```

However, when using different transformations in different layers we need to apply them in each layer. Here we compare three different smoothing methods.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line(data = . %>% smooth_spct(method = "custom"), 
            colour = "cornflowerblue", linewidth = 0.7) +
  geom_line(data = . %>% smooth_spct(method = "lowess"), 
            colour = "green", linewidth = 0.7) +
  geom_line(data = . %>% smooth_spct(method = "supsmu"), 
            colour = "red", linewidth = 0.7)
```

Of course, this approach works both with _geoms_ and _stats_, but one should remember that these layer functions do not "see" the original data objects (neither the default nor that passed as argument to the layers' `data` parameter), but instead a new `data.frame` containing the mapped variables in columns named according to _aesthetics_. The next example demonstrates this and illustrates that smoothing displaces the wavelength of maximum spectral irradiance.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line() + 
  stat_peaks(size = 3, span = NULL) +
  stat_peaks(geom = "vline", linetype = "dotted", span = NULL) +
  geom_line(data = . %>% smooth_spct(method = "supsmu"), 
            colour = "red", linewidth = 1) +
  stat_peaks(data = . %>% smooth_spct(method = "supsmu"),
             colour = "red", size = 3, span = NULL) +
  stat_peaks(data = . %>% smooth_spct(method = "supsmu"),
             geom = "vline", colour = "red", 
             linetype = "dotted", span = NULL)
```

We can easily highlight a wavelength range by over-plotting the same line in a different colour.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line() + 
  geom_line(data = . %>% trim_wl(range = PAR()), colour = "red")
```

We can highlight a range of wavelengths by plotting the points using colours matching human colour vision. Method `tag()` adds colour definitions as a new column named `wl.color` to the default `data` of the layer, after wavelengths outside the range of wavelengths of visible light have been dropped or "trimmed out" by method `trim_wl()`.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_line() + 
  geom_point(data = . %>% trim_wl(range = VIS()) %>% tag(),
            mapping = aes(color = wl.color),
            shape = "circle", size = 1.3) +
  scale_color_identity()
```

In the plot above, spectral irradiance as well as the wavelength at each data
point are taken into account when computing the colours, while below, only the
wavelength at the centre of each waveband is used because we passed 
`w.band = VIS_bands()` when calling `tag()`.

```{r, eval = eval_chunks}
ggplot(sun.spct) + 
  geom_area(data = . %>% trim_wl(range = VIS()) %>% tag(w.band = VIS_bands()),
            mapping = aes(fill = wb.color)) +
  geom_line() + 
  scale_fill_identity()
```

The examples above made use of `sun.spct` an object belonging to class `"source_spct"`, containing data for a single spectrum. Package 'photobiology' defines classes for different types of spectral data, and a base class `"generic_spct"`. Any object of belonging to one of these classes can be used as shown above for `sun.spct`. Next we show examples involving operations involving multiple spectra stored in a single R object.

Some of the methods from 'photobiology' are also defined for `data.frame` and can be used as summary functions with data that are not radiation spectra, such as any `data` _seen_ by layer functions including `stat_summary()`. Furthermore, on-the-fly summaries and transformation can be used in any ggplot layer function and with any suitable function accepting data frames as input.

### Multiple spectra

```{r, include=FALSE, echo=FALSE}
opts_chunk$set(fig.width = 7, fig.height = 7)
```

Package 'photobiology' supports two alternative ways of storing multiple spectra in a single R object: collections of spectra in objects of classes with names ending in `_mspct` such as `source_mspct` and in objects of classes with names ending in `_spct` such as `source_spct`
in objects. When passed to `ggplot()` they are both added to the `"gg"` object as objects of one
of the `_spct` classes. (The former are an specialization of lists of data frames, while the second are an specialization of data frame, and, thus, compatible with 'ggplot2'.)

We use for these examples data for a time series of five spectra measured close to sunset. Objects `sun_evening.spct` and
`sun_evening.mspct` contain the same data but belong to classes `"source_spct"`
and `"source_mspct"`, respectively.

We use the two spectra extracted from the collection `sun_evening.mspct`, as two spectra are enough to demonstrate how on-the-fly operations work with multiple spectra. This first example is just a plain plot with two panels.

```{r}
ggplot(data = sun_evening.mspct[c(2, 5)]) +
  aes(linetype = spct.idx) +
  geom_line() +
  facet_wrap(facets = vars(spct.idx), ncol = 1)
```

Similarly as shown above for a single spectrum, we can smooth multiple spectra.

```{r}
ggplot(data = sun_evening.mspct[c(2, 5)]) +
  aes(linetype = spct.idx) +
  geom_line(colour = "blue", alpha = 0.5) +
  geom_line(data = . %>% smooth_spct(method = "supsmu")) +
  facet_wrap(facets = vars(spct.idx), ncol = 1)
```

Normalizing the spectra to one at their maxima gives a clearer view of how they differ in shape. However, an annotation of the plot giving the irradiance before normalization can help understand the data.

```{r}
ggplot(data = sun_evening.mspct[c(2, 5)]) +
  aes(linetype = spct.idx) +
  geom_line(data = . %>% normalize()) +
  stat_wb_irrad(time.unit = "second",
                unit.in = "energy",
                w.band = list(PAR(), NIR()),
                label.fmt = "%.1#f~W~m^{-2}",
                parse = TRUE,
                colour = "black",
                size = 2.5) +
  facet_wrap(facets = vars(spct.idx), ncol = 1)
```

It can be even clearer to scale the spectra so that the areas under the different curves are the same.

```{r}
ggplot(data = sun_evening.mspct[c(2, 5)]) +
  aes(linetype = spct.idx) +
  geom_line(data = . %>% fscale()) +
  stat_wb_irrad(time.unit = "second",
                unit.in = "energy",
                w.band = list(PAR(), NIR()),
                label.fmt = "%.1#f~W~m^{-2}",
                parse = TRUE,
                colour = "black",
                size = 2.5) +
  facet_wrap(facets = vars(spct.idx), ncol = 1)
```

Transforming `data` when calling layer functions is useful, as shown here, when different transformations to the data are applied in different layers of the same plot.

While above we used two spectra extracted from `sun_evening.mspct`, here we use `sun_evening.spct`, and
plot all five spectra.

```{r}
ggplot(data = sun_evening.spct) +
  aes(linetype = spct.idx) +
  geom_line() +
  geom_line(data = . %>% smooth_spct(method = "supsmu"),
            colour = "red") +
  facet_wrap(facets = vars(spct.idx), ncol = 2)
```

Another use case is when a layer displays a summary spectrum computed from multiple spectra shown in a different plot layer. In the example below the blue line shows the median spectrum. Other spectra are semitransparent and show as grey lines. To
avoid the mapping to `group` for the plot as a whole we pass
`aes()` as mapping to be merged with the automatic mapping of
_x_ and _y_.

```{r, fig.height= 5}
ggplot(data = sun_evening.spct, aes()) +
   geom_line(aes(group = .data[[id_factor(sun_evening.spct)]]),
             alpha = 1/3) +
   geom_line(data = . %>% s_median(), # parallel median
             colour = "blue")
```

There are few situations where applying transformations to multiple
spectra does not result in overcrowded plots, one example being animated plots 
where different layers are displayed sequentially. Further examples of 
animations are available in the separate on-line only package article 
[Animated plots of spectral data](https://docs.r4photobiology.info/ggspectra/articles/animated-plots.html). In this case a smoothed version of each spectrum is displayed in red and the measured spectrum in black.

```{r, fig.height=5, eval=eval_gganimate}
ggplot(data = sun_evening.mspct) +
  geom_line() +
  geom_line(data = . %>% smooth_spct(method = "supsmu"),
            colour = "red") +
  transition_states(spct.idx,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Now showing {closest_state}',
          subtitle = 'Frame {frame} of {nframes}')
```
